---
title: "Mortality forecasting"
subtitle: "Vignette version number 0.2.0"
date: "Last reviewed 06/10/21"
author: "Duncan Gillespie"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mortality forecasting}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
bibliography: mort-refs.bib
header-includes:
  \usepackage{float}
  \usepackage{amsmath}
link-citations: yes
citation_package: natbib
biblio-style: vancouver
urlcolor: blue
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE, results = 'hide', warning = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.pos = 'H'
)

library(readxl)
library(data.table)
suppressPackageStartupMessages(library(ggplot2))
#library(cowplot)

#knitr::opts_knit$set(root.dir = "/Volumes/Shared/ScHARR/PR_STAPM/Code/R_packages/mort.tools")
#knitr::opts_knit$set(root.dir = "X:/ScHARR/PR_STAPM/Code/R_packages/mort.tools")

```


The mortality forecasting code developed and discussed here is still experimental. There are several potential methods that could be used to forecast mortality by cause, and further thought is needed about what the most reliable method might be.

The following code illustrates the observed and forecast cause and subgroup specific mortality rates that would be used in STAPM. In this case, we forecast the trends from 2016 to 2100.  

```{r forecast example code, eval = F}

#install.packages("X:/ScHARR/PR_STAPM/Code/R_packages/mort.tools_1.0.0.zip", 
  # repos = NULL, type = "binary")

library(mort.tools)
library(ggplot2)
library(readxl)

# The observed mortality rates to input into the smoothing and forecasting process
mort_data_cause <- fread("X:/ScHARR/PR_Mortality_data_TA/Code/model_inputs/Output/
                         tob_death_rates_national_2019-05-06_mort.tools_1.0.0.csv")

# The paramaters that control the smoothing and forecast methods for each cause
params <- setDT(read_xlsx("data-raw/Mortality data/
                          tobacco mortality forecasting parameters.xlsx"))

# Run the forecast
cforecast <- CombinedForecast(
  data = mort_data_cause,
  forecast_params = params,
  n_years = 2100 - 2016, # time horizon - jumpoff year
  folder_path = "data-raw/Mortality data"
)

tob_mort_data_cause <- copy(cforecast$mx_data_cause)

```

